---
title: "01_load"
format: html
---

# Load raw data

```{r}
#load libraries
library(readxl)
library(readr)
library(dplyr)
library(tidyr)
library(stringr)
library(purrr)
library(here)
```

```{r}
# Path to the raw Excel file
raw_file_path <- here("data", "_raw", "AP_cohort_COVID_1_raw.xlsx")
sheets <- excel_sheets(raw_file_path)

#Get a vector of sheet names (one per patient/sample)
sheet_list <- map(sheets, 
                  ~ read_xlsx(raw_file_path, 
                              sheet = .x))
names(sheet_list) <- sheets
```

------------------------------------------------------------------------

## Export each Excel sheet as a TSV

Here we read each sheet and save it as its own `.tsv` file.\

This is mainly for transparency / debugging: you can open a single TSV if something looks odd.

```{r}
#create tsv-files from nested tibble

# Directory where we will save all sheet-wise TSV files

out_dir <- here("data", "sheets_tsv")

# Create the directory if it doesn't already exist

dir.create(out_dir, showWarnings = FALSE)

# Read each sheet and write it to a .tsv with the same name as the sheet

sheet_list <- map(sheets, ~ read_xlsx(raw_file_path, sheet = .x))
names(sheet_list) <- sheets

# Write each sheet to its own TSV file

imap(
sheet_list,
function(df, sheet_name) {
out_file <- here(out_dir, paste0(sheet_name, ".tsv"))
write_tsv(df, out_file)
}
)

```

At this point we should have a folder data/sheets_tsv/ with one TSV per sheet.

------------------------------------------------------------------------

## Extract multimer frequencies from each sheet

The experimental multimer frequency (from FACS) is stored in column AG, row 7 on each sheet. We read that cell for each sheet, clean it, and put everything in one tibble.

We also guard against:

Sheets that do not have AG7 → we return NA for those.

```{r}
# Helper function to get multimer frequency from cell AG7 on a given sheet

get_multimer_freq <- function(sh) {

# Try to read cell AG7 from the sheet

cell_tbl <- tryCatch(
read_xlsx(
raw_file_path,
sheet = sh,
range = "AG7",
col_names = FALSE
),
error = function(e) {
# If AG7 doesn't exist for this sheet, return an empty tibble
tibble()
}
)

# If AG7 is outside the sheet (no columns), set NA

if (ncol(cell_tbl) == 0) {
mf_val <- NA_real_
} else {
raw_val <- cell_tbl |> pull(1)

# Remove percent signs and whitespace, then coerce to numeric
cleaned <- raw_val |>
  stringr::str_replace_all("%", "") |>
  stringr::str_trim()

mf_val <- suppressWarnings(as.numeric(cleaned))

}

tibble(
sheet = sh,
multimer_frequency = mf_val
)
}

# Apply to all sheets

multimer_freq_df <- map_dfr(sheets, get_multimer_freq)

multimer_freq_df

```

You should see one row per sheet, and NA only where no multimer frequency is available.

------------------------------------------------------------------------

## Load all TSVs and stack them (also filter columns)

Now we read all the exported TSV files back in and combine them into a single tibble.

We also:

-   Add a sheet column based on the file name.

-   Ensure we always have a sample column (if it's missing, we use the sheet name).

-   keep ONLY the columns we care about

```{r}
#List all .tsv files we just wrote
tsv_files <- list.files(out_dir, pattern = "\\.tsv$", full.names = TRUE)

#Helper: extract sheet name from file path (strip directory + .tsv)
get_sheet_name <- function(path) {
  basename(path) |> str_remove("\\.tsv$")
}

all_peptides <- map_dfr(tsv_files, function(f) {
  df <- read_tsv(f, show_col_types = FALSE)
  sheet_name <- get_sheet_name(f)
  
  df |>
    # make sure we always have a sample + sheet column
    mutate(
      sheet  = sheet_name,
      sample = if ("sample" %in% names(df)) sample else sheet_name
    ) |>
    
    # keep ONLY the columns you care about
    select(
      sheet,
      sample,
      any_of(c(
        "count.1",
        "input.1", "input.2", "input.3",
        "log_fold_change",
        "p",
        "HLA",
        "Protein",
        "Peptide"
      ))
    )
})

#quick look:
all_peptides
```

## Join multimer frequencies to peptide table

We now attach the per-sheet `multimer_frequency` to every peptide row via the `sheet` variable.

```{r}
peptides_with_mf <- all_peptides |>
  left_join(multimer_freq_df, by = "sheet")

```

------------------------------------------------------------------------

## Create helper variables: HLA_Peptide, Response, Est.frequency

We now create the derived variables needed for analysis:

-   `HLA_Peptide`: concatenation of HLA allele and peptide sequence.

-   `Response`: TRUE if

    -   `log_fold_change > 2`

    -   `p < 0.05.`

-   `Est.frequency`: estimated frequency of each responding peptide-specific T cell population within the total response for that sample:

    (Only defined for rows where `Response == TRUE`. Non-responders get `NA`.)

```{r}
peptides_final <- peptides_with_mf |>
  mutate(
    # HLA–peptide combo
    HLA_Peptide = paste(HLA, Peptide, sep = "_"),
    
    # Response flag: log_fold_change > 2 and p < 0.05
    Response = log_fold_change > 2 & p < 0.05,
    mutate(Response = if_else(Response, 1L, 0L))
  ) |>
  group_by(sample) |>
  mutate(
    # sum of count.1 only among responses within each sample
    total_count_responses = sum(count.1[Response], na.rm = TRUE),
    
    # Est.frequency = count.1 / sum(count.1 for responses) * multimer_frequency
    Est.frequency = if_else(
      Response &
        total_count_responses > 0 &
        !is.na(multimer_frequency),
      count.1 / total_count_responses * multimer_frequency,
      NA_real_
    )
  ) |>
  ungroup()

```

We expect:

-   `Response` mostly FALSE (most peptides do not respond).

-   `Est.frequency` is only non-NA for rows where Response == TRUE.

```{r}
# How many responses vs non-responses?

peptides_final |>
count(Response)

# Look at some responding peptides

peptides_final |>
filter(Response) |>
select(sample, HLA, Peptide, count.1, log_fold_change, p, multimer_frequency, Est.frequency) |>
head(20)

```

------------------------------------------------------------------------

## Save final combined dataset

Finally, we save the combined, augmented dataset as a TSV.\

We keep only the columns that are relevant for later steps.

```{r}
# Select and order columns for output

peptides_out <- peptides_final |>
select(
sample,
count.1,
input.1, input.2, input.3,
log_fold_change,
p,
HLA,
Protein,
Peptide,
HLA_Peptide,
Response,
Est.frequency
)

# Create a "clean" directory if needed

clean_dir <- here("data", "clean")
dir.create(clean_dir, showWarnings = FALSE)

# Write master peptide table

write_tsv(
peptides_out,
file = here("data", "clean", "peptides_master.tsv")
)


```
