---
title: "02_clean"
format: html
---

```{r}
# Read data/01_dat_load.tsv
# Clean and convert to tidy format
# Save as data/02_dat_clean.tsv
```

## 

## Load the libraries

```{r}
library("tidyverse")
library("here")
library("purrr")
library("readxl")

```

## Load the data

```{r}

path <- here("data", "sheets_tsv")
raw_data <- list.files(path, pattern = "tsv$",full.names = TRUE )

combined_raw <- map(raw_data, function(f) {
  read_tsv(f, col_types = cols(
    `-log10(masked_p)` = col_double(), 
    .default = col_guess()             
  ))
}) |>  
  list_rbind()

```

We kept getting the error that the column named "-log10(masked_p) " is sometimes a character and sometimes a numeric (double), so we forced to always be a numeric value.

Now we start filtering for the columns we are interested in

```{r}
combined_raw |> 
  colnames()
```

```{r}
subset_combined <- combined_raw |>
  select("barcode", "sample", "count.1", "input.1",  "input.2", "input.3", "log_fold_change", "p" , "HLA", "Protein", "Peptide")
```

This let's discuss where when we added but we need to add the multimer frequencies as well, maybe before and bind them together

## Load multimer frequencies

```{r}
multimers_frequencies <- read_excel(here("data", "_raw", "AP_cohort_COVID_1_multimer_frequency_raw.xlsx"))
```

Now we join the tables

```{r}
 join_all <- subset_combined |> 
  left_join(multimers_frequencies)
```

Now we add the columns that we are missing

```{r}
#HLA-peptide
join_all <- join_all |> 
  mutate(HLA_peptide = str_c(HLA, Peptide, sep = "-" ))

#response (we consider a response when it is a logfold change above 2)
join_all <- join_all |> 
  mutate(response = case_when(log_fold_change > 2 ~ 1,
                              log_fold_change <= 2 ~ 0))

#Estimated frequency : This data is calculated according to the following formula: count.1 / (sum of count.1 for RESPONSES) * Multimer Frequency.

join_all <-join_all |> 
  filter( response == 1) |> 
  group_by(sample) |> 
  mutate(sum_count.1 = sum(count.1)) |> 
  mutate(estimated_frequency = (count.1 /sum_count.1) * Multimer_positive ) |> 
  ungroup()

```

```{r}
#total estimated frequencies should be equal to multimer frequencies, it is a way to validate that what you did it correct
join_all <- join_all |> 
  group_by(sample) |> 
  mutate(total_estimated_frequency = sum(estimated_frequency)) |> 
  ungroup()
```

Lastly let's modify the patients' name removing unnecessary information

```{r}
x <- join_all |> 
  mutate(sampleID = str_remove(sample, "_.*$"),
         sampleID = str_sub(sampleID, start = 3, end = 4)) |> 
  relocate (sampleID)


x|> 
  select(sample) |> 
  unique()

x |> 
  select(sampleID) |> 
  unique()

  
```

Okay to discuss that 19 is there twice so in the renaming wecneed to do sometihng about it
