---
title: "02_clean"
format: html
---

```{r}
# Read data/01_dat_load.tsv
# Clean and convert to tidy format
# Save as data/02_dat_clean.tsv
```

```{r}
library("tidyverse")
library("here")
```

Create a combined tsv

```{r}
#| warning: false

file_path <- here("data", "sheets_tsv")

participants <- list.files(
  path = file_path,
  full.names = TRUE)



file_names <- basename(participants)
names(participants) <- file_names


#define correct interpretation of column which raised error
col_spec <- cols(
    `-log10(masked_p)` = col_double()
)


#vertically stacking as each patient has same cols
combined_data <- participants |>
  map_dfr(
    .f = ~ read_tsv(
        file = .x,
        col_types = col_spec, 
        na = c("NA", "N/A", "---")
    ),
    .id = "patient_sample"
  )

combined_data
```

Extract PE+ Population = Multimer Frequency

```{r}
combined_data <- combined_data |>
  group_by(sample)|>
  rename(freq= ...33)|>
  mutate(
    MulitmerFrequency = sum(replace_na(freq, 0))
  )|>
  ungroup()


combined_data
```

Drop last row (only contains summary statistics)

```{r}
combined_data <- combined_data |>
  slice(-n())
```

Keep only important columns

```{r}
subset_data <-combined_data |>
  select(
    c("sample", 
      "count.1", 
      "input.1", 
      "input.2", 
      "input.3", 
      "log_fold_change", 
      "p", 
      "HLA", 
      "Protein", 
      "Peptide",
      "MulitmerFrequency")
  )
```

Extract patient number

```{r}
combined_data_new <- subset_data |>
  separate(
    col = sample,
    into = c("sample", "ending"),
    sep = "_"  )|>
  select(-ending)|>
  mutate(
    #Regex pattern: ^ start
    # [A-Z]: uppercaseLetters
    #(): what to extract
    # \d: digits, {2}: exactly 2
    #\\d two more digits in name
    #$ end of string
    patient_id = str_match(sample, "^[A-Z]+(\\d{2})\\d{2}$")[, 2]
  )|>
  select(
    last_col(),
    everything() 
  )


combined_data_new
```

IS THIS ALREADY WRANGLING???

Create new columns

HLA-peptides (HLA-peptide combination)

```{r}
combined_data_new <- combined_data_new |>
  mutate(
    'HLA-peptide' = paste(HLA, Peptide)
  )
combined_data_new
```

Identify response by log_fold_change and p-value

```{r}
combined_data_new <- combined_data_new |>
  mutate(
    response = if_else(
      (log_fold_change > 2.0 & p < 0.05),
      1,
      0)
    )
combined_data_new
```

Estimated Frequency (based on the authors:

-   count.1 / (sum of count.1 for Responses) \* multimer Frequency

```{r}
complete_data <- combined_data_new |>
  group_by(patient_id)|>
  mutate(
    ResponseCount = sum(if_else(response == 1,
                                count.1, 
                                0
                                )
                        )
    )|>
  ungroup()|>
  mutate(
    estimatedFrequency = case_when(
      ResponseCount == 0 ~ 0,
      response == 1 ~ (replace_na(count.1, 0) / ResponseCount) * MulitmerFrequency,
      TRUE ~ 0
    )
      
  )

complete_data
```
