---
#title: "Analysis"
format: html
---

# 4. Analysis

------------------------------------------------------------------------

Script: **04_analysis.qmd**

**Purpose**: Analyse the how the peptides are distributed over HLA, Protein and patients

**Input**: 03_dat_aug.tsv

------------------------------------------------------------------------

```{r}
#| warning: false
#| message: false

# Load libraries ------------------------------------------------------------
library("here")
library("tidyverse")
library("patchwork")
```

```{r}
#| warning: false
#| message: false

# Load helper functions -----------------------------------------------------
source("99_proj_func.R")
```

```{r}
# Load augmented data -----------------------------------------------------
experiment_data <- load_tsvData("03_data_aug.tsv", show_col_types = FALSE)
```

## Manhattan plot: Immunogenic peptides, sorted by HLAs

```{r}
#| warning: false
#| fig.width: 11
#| fig.height: 6
#| dpi: 200

protein_colors <- c(
  "ORF1" = "#009E73",
  "S" = "#56B4E9",  
   "ORF3"= "#D55E00",
  "ORF6" = "#CC79A7",
  "ORF7" = "#00AFAA",
  "ORF8" = "#F0E442",
  "ORF10" = "#0099BB",
  "E" = "#0072B2",
  "M" =  "#E69F00",
  "N" = "#5A61FF"
)


manhattan_HLA <- experiment_data |>
  mutate(
    point_color = if_else(response == 0, "grey",protein_colors [Protein]),     
    point_fill  = if_else(response == 0, "grey", protein_colors [Protein]),
    HLA = factor(HLA, levels = c("A01:01","A02:01","A03:01","A24:02", "B07:02", "B08:01", "B15:01"))) 

  
  
manhattan_HLA |> 
    ggplot(aes(
    x = `HLA-peptide`,
    y = log_fold_change,
    colour = point_color,
    fill   = point_fill,
    size = estimated_frequency
  )) +
  geom_point() +

  facet_grid(~ HLA, scales = "free_x", space = "free") +
  geom_hline(yintercept = 2, linetype = "dashed") +
  scale_color_identity(guide = "none") +   
  scale_fill_identity(guide = "none") +
  scale_y_continuous(limits = c(0, NA), breaks = seq(0, 14, by = 2)) +
  #scale_y_break(breaks = c(0, 2)) +
 theme(
  panel.spacing = unit(0, "lines"),  
  panel.border = element_rect(colour = "grey", fill = NA, size = 0.3),
  strip.text = element_text(angle = 90, hjust = 0.5, size = 10, face = "bold"),
  legend.position = "bottom",
   axis.text.y.right = element_blank(),
  axis.ticks.y.right = element_blank(),
  strip.background = element_blank(),
  axis.text.x = element_blank(),
  axis.text.y.left = element_text(size = 12),
  title = element_text(size = 14),
    legend.title = element_text(size = 10)

   ) +

  labs(title = "Manhattan plot of T cell responses against SARS-CoV-2 antigens, sorted by HLA",
       x = "Peptide HLA pairs",
       y = "Log fold change",
       size = "Estimated frequency"
) 

```

In this plot, we can appreciate the distribution of immunogenic peptides across different HLAs, by displaying the estimated frequency of each peptide-HLA pair in the patient cohort.

## Summary bar plot: Peptides distribution across HLAs

```{r}

#how many peptides per HLA for positive T-responses
summary_HLA <- 
  experiment_data|>
  mutate(HLA = str_remove(HLA, ":"))|>
  filter(response ==1)|>
  group_by(HLA)|>
  summarize(peptidesPerHLA = n_distinct(Peptide))|>
  ungroup()|>
  #order HLA descending + fixes appearance for plotting
  arrange(desc(HLA)) |> 
  mutate(HLA = fct_inorder(HLA))

#compute all peptides per HLA independent of positive/negative T-responses
totalPeptidesPerHLA <- experiment_data|>
  group_by(HLA)|>
  summarize(totalPeptidesPerHLA = n_distinct(Peptide)) |> 
  arrange(desc(HLA)) |> 
  #also need the same order for combining correct values
  mutate(HLA = fct_inorder(HLA)) |>
  pull(totalPeptidesPerHLA)




pl1 <- summary_HLA |>
  ggplot(
    aes(x= peptidesPerHLA, 
        y = HLA)
  ) + 
  geom_col(fill = "darkblue")+
  geom_text(aes(label = peptidesPerHLA), hjust = -0.2)+
  scale_x_continuous(expand = expansion(mult = c(0,.2)))+
  labs(caption = "Number of HLA-specific\nSARS-CoV-2 epitopes")+
  theme(aspect.ratio = 2/1, 
        axis.title.x = element_blank(), 
        axis.title.y = element_blank(), 
        axis.ticks = element_line(color = "black"),
        axis.line = element_line(color = "black"),    
        panel.background = element_rect(fill = "white"),
        plot.caption = element_text(size = 12)
  )

pl2 <- summary_HLA |>
  mutate(
    percentage_peptides = peptidesPerHLA/totalPeptidesPerHLA * 100
  )|>

   ggplot(
    aes(x= percentage_peptides, 
        y = HLA)
  ) + 
  geom_col(fill = "darkblue")+
  geom_text(aes(label = round(percentage_peptides, digits = 1)), 
            hjust = -0.2)+
  scale_x_continuous(expand = expansion(mult = c(0,.2)))+
  labs(caption = "Immunogenic peptides\n (%, normalized)")+
  theme(aspect.ratio = 2/1, 
        axis.title.x = element_blank(), 
        axis.title.y = element_blank(), 
        axis.text.y  = element_blank(),
        axis.ticks = element_line(color = "black"),
        axis.line = element_line(color = "black"),    
        panel.background = element_rect(fill = "white"),
    plot.caption = element_text(size = 12)
  )
  


combined_HLAs <- pl1 | pl2 +
  plot_layout(widths = c())

combined_HLAs
```

The plot compares, for each HLA, the total number of distinct SARS-CoV-2 peptides that had a positive T-cell response. The right panel shows the percentage of immunogenic peptides normalized to the total peptide number of each HLA.\
This highlights the absolute and relative immunogenicity of different HLA types.

## Manhattan plot: Immunogenic peptides, sorted by protein of origin

```{r}
#| warning: false
#| fig.width: 11
#| fig.height: 6
#| dpi: 200

manhattan_protein <- experiment_data |>
  mutate(
    point_color = if_else(response == 0, "grey",protein_colors [Protein]),     
    point_fill  = if_else(response == 0, "grey", protein_colors [Protein]),
    Protein = factor(Protein, levels = c("ORF1","S","ORF3","N", "M", "ORF8", "ORF7", "ORF6","ORF10", "E")))
  
  
manhattan_protein |> 
    ggplot(aes(
    x = `HLA-peptide`,
    y = log_fold_change,
    colour = point_color,
    fill   = point_fill,
    size = estimated_frequency
  )) +
  geom_point() +

  facet_grid(~ Protein, scales = "free_x", space = "free") +
  geom_hline(yintercept = 2, linetype = "dashed") +
  scale_color_identity(guide = "none") +   
  scale_fill_identity(guide = "none") +
  scale_y_continuous(limits = c(0, NA), breaks = seq(0, 14, by = 2)) +
  #scale_y_break(breaks = c(0, 2)) +
 theme(
  panel.spacing = unit(0, "lines"),  
  panel.border = element_rect(colour = "grey", fill = NA, size = 0.3),
  strip.text = element_text(angle = 90, hjust = 0.5, size = 10, face = "bold"),
  legend.position = "bottom",
   axis.text.y.right = element_blank(),
  axis.ticks.y.right = element_blank(),
  strip.background = element_blank(),
  axis.text.x = element_blank(),
  axis.text.y.left = element_text(size = 12),
  title = element_text(size = 14),
    legend.title = element_text(size = 10)

   ) +

  labs(title = "Manhattan plot of T cell responses against SARS-CoV-2 peptides across patients, by protein",
       x = "Peptide HLA pairs",
       y = "Log fold change",
       size = "Estimated frequency"
) 

```

This plot gives an overview of how immunogenic peptides are spread across various proteins of origin, keeping into account their estimated frequency in the patient cohort.

## Summary bar plot: peptides distribution across protein of origin

```{r}
summary_protein <- experiment_data|>
  group_by(Protein)|>
      summarize(total = n_distinct(`HLA-peptide`),
                number_Tcells = sum(response == 1))


pl1 <- summary_protein |>
  ggplot(
    aes(x = number_Tcells,
        y = Protein,
        fill = Protein)
  )+
  geom_col()+
  scale_fill_manual(values  = protein_colors) +
  geom_text(aes(label = number_Tcells),hjust = -0.2)+
  scale_x_continuous(expand = expansion(mult = c(0,.2)))+
  labs(caption = "SARS-CoV-2\nT cell responses")+
  theme(aspect.ratio = 2/1, 
        axis.title.x = element_blank(), 
        axis.title.y = element_blank(), 
        axis.ticks = element_line(color = "black"),
        axis.line = element_line(color = "black"),    
        panel.background = element_rect(fill = "white"),
        plot.caption = element_text(size = 12),
        legend.position = "none"
  )


pl2 <- summary_protein |>
  mutate(
    percentage_peptides = number_Tcells/total * 100
  )|>
  ggplot(
    aes(x = percentage_peptides,
        y = Protein,
        fill = Protein)
  )+
  geom_col()+
  scale_fill_manual(values  = protein_colors) +
  geom_text(aes(label = round(percentage_peptides, digits = 1)),hjust = -0.2)+
  scale_x_continuous(expand = expansion(mult = c(0,.2)))+
  labs(caption = "Immunogenic peptides\n(%, normalized)")+
  theme(aspect.ratio = 2/1, 
        axis.title.x = element_blank(), 
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_line(color = "black"),
        axis.line = element_line(color = "black"),    
        panel.background = element_rect(fill = "white"),
        plot.caption = element_text(size = 12),
        legend.position = "none"
  )
  
combined_protein <- pl1 | pl2 +
  plot_layout(widths = c())

combined_protein
```

The plot compares, for each SARS-CoV-2 protein, the total number of HLA-peptide combination that resulted in a positive T-cell response. The right panel shows the percentage of immunogenic peptides relative to all detected peptides of that protein.\
This highlights the absolute and relative immunogenicity of different proteins.

## Patient-based distributions of peptides across proteins of origin

It is interesting to assess how different genes from SARS-CoV-2 genome contribute in generating immunogenic epitopes, for each patient in the cohort. To do this, we will plot the total frequency of T cell responses for each patient, distributed across the whole genome.

```{r}
#| warning: false
#| fig.width: 11
#| fig.height: 6
#| dpi: 200



patient_protein <- experiment_data |>
  mutate(estimated_frequency = estimated_frequency*100, #converts into percentage
         total_estimated_frequency = total_estimated_frequency*100,
         sampleID = fct_reorder(sampleID, total_estimated_frequency)) |>
  arrange(desc(total_estimated_frequency))

patient_protein|>
ggplot(aes(x = estimated_frequency, 
           y = sampleID, 
           fill = Protein)) +
  geom_col( position = "stack") +
  scale_fill_manual(values  = protein_colors) +
  scale_x_continuous(breaks= c(0, 5, 10, 15, 20, 25)) +
  labs( x = "SARS-CoV-2 reactive CD8 T cells (%)", 
        y = "Samples", 
        title = "Frequency of SARS-CoV-2 reactive CD8 T cells in patients with COVID19, distributed across protein of origin")
```

As shown here, for most samples, the majority of reactive T cells recognize antigens derived from ORF1. Taking into account the size of the genes, hence the amount of peptides that can be generated, it is reasonable to witness a bigger response against bigger genes (such as ORF1).

## Identifying immunodominant peptides

At last, we are interested in identifying which peptides are conserved across patients, keeping in mind the restrictions guided by HLA haplotypes. We will define immunodominant, those peptide-HLA for which a T cell population was identified in \>= 50 % of the patients carrying given HLA.

### Concept of Immunodominance

Immunodominance refers to the phenomenon where the immune system focuses its response on a small fraction of the available viral peptides. Even though SARS-CoV-2 has thousands of potential epitopes, only a select few trigger a strong T-cell response in the majority of patients.

Identifying these "immunodominant" epitopes is crucial for **vaccine design**, as a good vaccine must include these key targets to ensure broad protection across the population (especially considering different HLA types).

### Visualization

The bar plot below shows the **Prevalence** of the top immunodominant peptides within their respective HLA groups.

```{r}
#| warning: false
#| fig.width: 10
#| fig.height: 8
#| dpi: 200

# --- Calculation ---
# 1. Calculate Prevalence per HLA
hla_counts <- experiment_data |> 
  distinct(sampleID, HLA) |> 
  count(HLA, name = "n_patients")

valid_hlas <- hla_counts |> 
  filter(n_patients >= 3) |> 
  pull(HLA)

prevalence_data <- experiment_data |> 
  filter(HLA %in% valid_hlas) |> 
  group_by(HLA, Peptide, Protein) |> 
  summarise(
    total_tested = n_distinct(sampleID),
    responders = sum(response),
    prevalence = responders / total_tested,
    .groups = "drop"
  ) |> 
  filter(prevalence > 0)

# 2. Filter for TOP immunodominant peptides
top_peptides <- prevalence_data |> 
  arrange(desc(prevalence)) |> 
  slice_head(n = 20) # Take top 20 strongest responses

# 3. Barplot
ggplot(top_peptides, aes(x = prevalence, y = reorder(Peptide, prevalence), fill = Protein)) +
  geom_col() +
  scale_fill_manual(values = protein_colors) +
  geom_vline(xintercept = 0.5, linetype = 2) +
  scale_x_continuous(labels = scales::percent, expand = c(0, 0.05)) +
    labs(
    title = "Top Immunodominant SARS-CoV-2 Epitopes",
    subtitle = "Peptides with highest response frequency within HLA groups (Top 20 shown)",
    x = "Prevalence (% of responders)",
    y = "Peptide",
    fill = "Protein"
  ) +
  theme(
    legend.position = "right",
    axis.text.y = element_text(size = 8)
  )
```
