---
title: "05_analysis_1"
format: html
---

```{r}
# First main analysis
# e.g. immunodominance landscape similar to Fig. 1E
```

Libraries

```{r}
library("here")
library("tidyverse")
```

```{r}
source("99_proj_func.R")
```

Load data

```{r}
experiment_data <- load_tsvData("03_data_augment.tsv", show_col_types = FALSE)
```

# Plot 1E

```{r}

```

# Plot 1F - Identified HLA-specific SARS-CoV-2 epitopes and HLA-restricted immunogenicity

```{r}
library("patchwork")
```

```{r}

#how many peptides per HLA for positive T-responses
extract_data <- 
  experiment_data|>
  mutate(HLA = str_remove(HLA, ":"))|>
  filter(response ==1)|>
  group_by(HLA)|>
  summarize(peptidesPerHLA = n_distinct(Peptide))|>
  ungroup()|>
  #order HLA descending + fixes appearance for plotting
  arrange(desc(HLA)) |> 
  mutate(HLA = fct_inorder(HLA))

#compute all peptides per HLA independent of positive/negative T-responses
totalPeptidesPerHLA <- experiment_data|>
  group_by(HLA)|>
  summarize(totalPeptidesPerHLA = n_distinct(Peptide)) |> 
  arrange(desc(HLA)) |> 
  #also need the same order for combining correct values
  mutate(HLA = fct_inorder(HLA)) |>
  pull(totalPeptidesPerHLA)




pl1 <- extract_data |>
  ggplot(
    aes(x= peptidesPerHLA, 
        y = HLA)
  ) + 
  geom_col(aes(width = 0.5), 
           fill = "darkblue")+
  geom_text(aes(label = peptidesPerHLA), hjust = -0.2)+
  scale_x_continuous(expand = expansion(mult = c(0,.2)))+
  labs(caption = "Number of HLA-specific\nSARS-CoV-2 epitopes")+
  theme(aspect.ratio = 2/1, 
        axis.title.x = element_blank(), 
        axis.title.y = element_blank(), 
        axis.ticks = element_line(color = "black"),
        axis.line = element_line(color = "black"),    
        panel.background = element_rect(fill = "white"),
        plot.caption = element_text(size = 12)
  )

pl2 <- extract_data |>
  mutate(
    percentage_peptides = peptidesPerHLA/totalPeptidesPerHLA * 100
  )|>

   ggplot(
    aes(x= percentage_peptides, 
        y = HLA)
  ) + 
  geom_col(aes(width = 0.5), 
           fill = "darkblue")+
  geom_text(aes(label = round(percentage_peptides, digits = 1)), 
            hjust = -0.2)+
  scale_x_continuous(expand = expansion(mult = c(0,.2)))+
  labs(caption = "Immunogenic peptides\n (%, normalized)")+
  theme(aspect.ratio = 2/1, 
        axis.title.x = element_blank(), 
        axis.title.y = element_blank(), 
        axis.text.y  = element_blank(),
        axis.ticks = element_line(color = "black"),
        axis.line = element_line(color = "black"),    
        panel.background = element_rect(fill = "white"),
    plot.caption = element_text(size = 12)
  )
  


combined <- pl1 | pl2 +
  plot_layout(widths = c())

combined
```

The plot compares, for each HLA, the total number of distinct SARS-CoV-2 peptides that resulted in a positive T-cell response. The right panel shows the percentage of immunogenic peptides normalized to the total peptide number of each HLA. \
This highlights the absolute and relative immunogenicity of different HLA types.

# Plot 1G

```{r}

```

# Plot 1H - Epitopes derived from each SARS-CoV protein and their immunogenicity score

```{r}
extract_data <- experiment_data|>
  group_by(Protein)|>
      summarize(total = n_distinct(`HLA-peptide`),
                number_Tcells = sum(response == 1))


pl1 <- extract_data |>
  ggplot(
    aes(x = number_Tcells,
        y = Protein,
        fill = Protein)
  )+
  geom_col(aes(width = 0.5))+
  geom_text(aes(label = number_Tcells),hjust = -0.2)+
  scale_x_continuous(expand = expansion(mult = c(0,.2)))+
  labs(caption = "SARS-CoV-2\nT cell responses")+
  theme(aspect.ratio = 2/1, 
        axis.title.x = element_blank(), 
        axis.title.y = element_blank(), 
        axis.ticks = element_line(color = "black"),
        axis.line = element_line(color = "black"),    
        panel.background = element_rect(fill = "white"),
        plot.caption = element_text(size = 12),
        legend.position = "none"
  )


pl2 <- extract_data |>
  mutate(
    percentage_peptides = number_Tcells/total * 100
  )|>
  ggplot(
    aes(x = percentage_peptides,
        y = Protein,
        fill = Protein)
  )+
  geom_col(aes(width = 0.5))+
  geom_text(aes(label = round(percentage_peptides, digits = 1)),hjust = -0.2)+
  scale_x_continuous(expand = expansion(mult = c(0,.2)))+
  labs(caption = "Immunogenic peptides\n(%, normalized)")+
  theme(aspect.ratio = 2/1, 
        axis.title.x = element_blank(), 
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_line(color = "black"),
        axis.line = element_line(color = "black"),    
        panel.background = element_rect(fill = "white"),
        plot.caption = element_text(size = 12),
        legend.position = "none"
  )
  
combined <- pl1 | pl2 +
  plot_layout(widths = c())

combined
```

The plot compares, for each SARS-CoV-2 protein, the total number of HLA-peptide combination that resulted in a positive T-cell response. The right panel shows the percentage of immunogenic peptides relative to all detected peptides of that protein. \
This highlights the absolute and relative immunogenicity of different protein types.

# Plot 1I

```{r}

```

# Plot 2A - funny plot

```{r}

```
