---
title: "05_analysis_1"
format: html
---

```{r}
# First main analysis
# e.g. immunodominance landscape similar to Fig. 1E
```

## Load the libraries

```{r}
library(dplyr) # better load tidyverse, already included there
library(ggplot2) # better load tidyverse, already included there
#library(ggbreak)
#library(ggforce)

```

```{r}
library("here")
library("tidyverse")
library("patchwork")
```

```{r}
source("99_proj_func.R")
```

## Load data

```{r}
experiment_data <- load_tsvData("03_data_augment.tsv", show_col_types = FALSE)
```

# Plot 1E

```{r}
#goodone

#we need to do it in clean
complete_data <- experiment_data |>
  mutate(Protein = str_to_upper(Protein))
  

protein_colors <- c(
  "ORF1" = "#1f77b4",
  "S" = "#ff7f0e",  
   "ORF3"= "#2ca02c",
  "ORF6" = "pink",
  "ORF7" = "red",
  "ORF8" = "lightblue",
  "ORF10" = "yellow",
  "E" = "purple",
  "M" = "darkblue",
  "N" = "darkred"
)

extras <- complete_data |>
  mutate(
    point_color = if_else(response == 0, "grey",protein_colors [Protein]),     
    point_fill  = if_else(response == 0, "grey", protein_colors [Protein]),
    HLA = factor(HLA, levels = c("A01:01","A02:01","A03:01","A24:02", "B07:02", "B08:01", "B15:01")))
  
  
  extras |> 
    ggplot(aes(
    x = `HLA-peptide`,
    y = log_fold_change,
    colour = point_color,
    fill   = point_fill,
    size = estimatedFrequency
  )) +
  geom_point() +

  facet_grid(~ HLA, scales = "free_x", space = "free") +
  geom_hline(yintercept = 2, linetype = "dashed") +
  scale_color_identity(guide = "none") +   
  scale_fill_identity(guide = "none") +
  scale_y_continuous(limits = c(0, NA), breaks = seq(0, 14, by = 2)) +
  #scale_y_break(breaks = c(0, 2)) +
 theme(
  panel.spacing = unit(0, "lines"),  
  panel.border = element_rect(colour = "grey", fill = NA, size = 0.3),
  strip.text = element_text(angle = 90, hjust = 0.5, size = 10, face = "bold"),
  legend.position = "bottom",
   axis.text.y.right = element_blank(),
  axis.ticks.y.right = element_blank(),
  strip.background = element_blank(),
  axis.text.x = element_blank(),
  axis.text.y.left = element_text(size = 12),
  title = element_text(size = 16),
    legend.title = element_text(size = 10)

   ) +

  labs(title = "HLA",
       x = "Peptide HLA pairs",
       y = "Log fold change",
       size = "Estimated frequency"
) 

```

# Plot 1F - Identified HLA-specific SARS-CoV-2 epitopes and HLA-restricted immunogenicity

```{r}

#how many peptides per HLA for positive T-responses
extract_data <- 
  experiment_data|>
  mutate(HLA = str_remove(HLA, ":"))|>
  filter(response ==1)|>
  group_by(HLA)|>
  summarize(peptidesPerHLA = n_distinct(Peptide))|>
  ungroup()|>
  #order HLA descending + fixes appearance for plotting
  arrange(desc(HLA)) |> 
  mutate(HLA = fct_inorder(HLA))

#compute all peptides per HLA independent of positive/negative T-responses
totalPeptidesPerHLA <- experiment_data|>
  group_by(HLA)|>
  summarize(totalPeptidesPerHLA = n_distinct(Peptide)) |> 
  arrange(desc(HLA)) |> 
  #also need the same order for combining correct values
  mutate(HLA = fct_inorder(HLA)) |>
  pull(totalPeptidesPerHLA)




pl1 <- extract_data |>
  ggplot(
    aes(x= peptidesPerHLA, 
        y = HLA)
  ) + 
  geom_col(aes(width = 0.5), 
           fill = "darkblue")+
  geom_text(aes(label = peptidesPerHLA), hjust = -0.2)+
  scale_x_continuous(expand = expansion(mult = c(0,.2)))+
  labs(caption = "Number of HLA-specific\nSARS-CoV-2 epitopes")+
  theme(aspect.ratio = 2/1, 
        axis.title.x = element_blank(), 
        axis.title.y = element_blank(), 
        axis.ticks = element_line(color = "black"),
        axis.line = element_line(color = "black"),    
        panel.background = element_rect(fill = "white"),
        plot.caption = element_text(size = 12)
  )

pl2 <- extract_data |>
  mutate(
    percentage_peptides = peptidesPerHLA/totalPeptidesPerHLA * 100
  )|>

   ggplot(
    aes(x= percentage_peptides, 
        y = HLA)
  ) + 
  geom_col(aes(width = 0.5), 
           fill = "darkblue")+
  geom_text(aes(label = round(percentage_peptides, digits = 1)), 
            hjust = -0.2)+
  scale_x_continuous(expand = expansion(mult = c(0,.2)))+
  labs(caption = "Immunogenic peptides\n (%, normalized)")+
  theme(aspect.ratio = 2/1, 
        axis.title.x = element_blank(), 
        axis.title.y = element_blank(), 
        axis.text.y  = element_blank(),
        axis.ticks = element_line(color = "black"),
        axis.line = element_line(color = "black"),    
        panel.background = element_rect(fill = "white"),
    plot.caption = element_text(size = 12)
  )
  


combined <- pl1 | pl2 +
  plot_layout(widths = c())

combined
```

The plot compares, for each HLA, the total number of distinct SARS-CoV-2 peptides that resulted in a positive T-cell response. The right panel shows the percentage of immunogenic peptides normalized to the total peptide number of each HLA.\
This highlights the absolute and relative immunogenicity of different HLA types.

# Plot 1G

```{r}
#goodone

#we need to do it in clean
complete_data <- experiment_data |>
  mutate(Protein = str_to_upper(Protein))
  

protein_colors <- c(
  "ORF1" = "#1f77b4",
  "S" = "#ff7f0e",  
   "ORF3"= "#2ca02c",
  "ORF6" = "pink",
  "ORF7" = "red",
  "ORF8" = "lightblue",
  "ORF10" = "yellow",
  "E" = "purple",
  "M" = "darkblue",
  "N" = "darkred"
)

extras <- complete_data |>
  mutate(
    point_color = if_else(response == 0, "grey",protein_colors [Protein]),     
    point_fill  = if_else(response == 0, "grey", protein_colors [Protein]),
    HLA = factor(HLA, levels = c("ORF1","S","ORF3","N", "M", "ORF8", "ORF7", "ORF6","ORF10", "E")))
  
  
  extras |> 
    ggplot(aes(
    x = `HLA-peptide`,
    y = log_fold_change,
    colour = point_color,
    fill   = point_fill,
    size = estimatedFrequency
  )) +
  geom_point() +

  facet_grid(~ Protein, scales = "free_x", space = "free") +
  geom_hline(yintercept = 2, linetype = "dashed") +
  scale_color_identity(guide = "none") +   
  scale_fill_identity(guide = "none") +
  scale_y_continuous(limits = c(0, NA), breaks = seq(0, 14, by = 2)) +
  #scale_y_break(breaks = c(0, 2)) +
 theme(
  panel.spacing = unit(0, "lines"),  
  panel.border = element_rect(colour = "grey", fill = NA, size = 0.3),
  strip.text = element_text(angle = 90, hjust = 0.5, size = 10, face = "bold"),
  legend.position = "bottom",
   axis.text.y.right = element_blank(),
  axis.ticks.y.right = element_blank(),
  strip.background = element_blank(),
  axis.text.x = element_blank(),
  axis.text.y.left = element_text(size = 12),
  title = element_text(size = 16),
    legend.title = element_text(size = 10)

   ) +

  labs(title = "Protein",
       x = "Peptide HLA pairs",
       y = "Log fold change",
       size = "Estimated frequency"
) 

```

# Plot 1H - Epitopes derived from each SARS-CoV protein and their immunogenicity score

```{r}
extract_data <- experiment_data|>
  group_by(Protein)|>
      summarize(total = n_distinct(`HLA-peptide`),
                number_Tcells = sum(response == 1))


pl1 <- extract_data |>
  ggplot(
    aes(x = number_Tcells,
        y = Protein,
        fill = Protein)
  )+
  geom_col(aes(width = 0.5))+
  geom_text(aes(label = number_Tcells),hjust = -0.2)+
  scale_x_continuous(expand = expansion(mult = c(0,.2)))+
  labs(caption = "SARS-CoV-2\nT cell responses")+
  theme(aspect.ratio = 2/1, 
        axis.title.x = element_blank(), 
        axis.title.y = element_blank(), 
        axis.ticks = element_line(color = "black"),
        axis.line = element_line(color = "black"),    
        panel.background = element_rect(fill = "white"),
        plot.caption = element_text(size = 12),
        legend.position = "none"
  )


pl2 <- extract_data |>
  mutate(
    percentage_peptides = number_Tcells/total * 100
  )|>
  ggplot(
    aes(x = percentage_peptides,
        y = Protein,
        fill = Protein)
  )+
  geom_col(aes(width = 0.5))+
  geom_text(aes(label = round(percentage_peptides, digits = 1)),hjust = -0.2)+
  scale_x_continuous(expand = expansion(mult = c(0,.2)))+
  labs(caption = "Immunogenic peptides\n(%, normalized)")+
  theme(aspect.ratio = 2/1, 
        axis.title.x = element_blank(), 
        axis.title.y = element_blank(),
        axis.text.y = element_blank(),
        axis.ticks = element_line(color = "black"),
        axis.line = element_line(color = "black"),    
        panel.background = element_rect(fill = "white"),
        plot.caption = element_text(size = 12),
        legend.position = "none"
  )
  
combined <- pl1 | pl2 +
  plot_layout(widths = c())

combined
```

The plot compares, for each SARS-CoV-2 protein, the total number of HLA-peptide combination that resulted in a positive T-cell response. The right panel shows the percentage of immunogenic peptides relative to all detected peptides of that protein.\
This highlights the absolute and relative immunogenicity of different protein types.

# Plot 1I- frequency of epitopes across proteins of origin

It is interesting to assess how different genes from SARS-CoV-2 genome contribute in generating immunogenic epitopes. To do this, we will plot the total frequency of T cell responses for each patient, distributed across the whole genome.

```{r}
augmented_data <- experiment_data |>
  mutate(estimatedFrequency = estimatedFrequency*100, #converts into percentage
         total_estimated_frequency = total_estimated_frequency*100,
         sampleID = fct_reorder(sampleID, total_estimated_frequency)) |>
  arrange(desc(total_estimated_frequency))



palette10 <- c(
  "#E69F00", "#56B4E9", "#009E73","#00AFAA", "#0072B2","#D55E00", "#CC79A7", "#F0E442","#0099BB", "#5A61FF")



augmented_data|>
ggplot(aes(x = estimatedFrequency, 
           y = sampleID, 
           fill = Protein)) +
  geom_col( position = "stack") +
  scale_fill_manual(values  = palette10) +
  scale_x_continuous(breaks= c(0, 5, 10, 15, 20, 25)) +
  labs( x = "SARS-CoV-2 reactive CD8 T cells (%)", 
        y = "Samples", 
        title = "Frequency of SARS-CoV-2 reactive CD8 T cells in patients with COVID19, distributed across protein of origin")
```

As shown here, for most samples, the majority of reactive T cells recognize antigens derived from ORF1. Taking into account the size of the genes, hence the amount of peptides that can be generated, it is reasonable to witness a bigger response against bigger genes (such as ORF1). However, we can perform some normalization to calculate the percentage of immunogenic peptides that arise from each gene. (Selina's figures).

# Plot 2A - funny plot

```{r}

```
