---
title: "05_analysis_1"
format: html
---

```{r}
# First main analysis
# e.g. immunodominance landscape similar to Fig. 1E
```

Libraries

```{r}
library("here")
library("tidyverse")
```

```{r}
source("99_proj_func.R")
```

Load data

```{r}
experiment_data <- load_tsvData("03_data_augment.tsv", show_col_types = FALSE)
```

# Plot 1E

```{r}

```

# Plot 1F

```{r}
library("patchwork")
```

WHAT normalization method should I use?? here: HLA(significant) / HLA ???

```{r}
experiment_data|>
  filter(HLA == "C07:02")
```

Plot 1: how many Peptides per HLA for positive T-cell responses

```{r}
#order the HLA in descending order
library(forcats)
```

```{r}
experiment_data |>
  mutate(HLA = str_remove(HLA, ":"))|>
  filter(response == 1)|>
  group_by(HLA)|>
  summarize(number = n_distinct(Peptide))|>
  ungroup()|>
  mutate(HLA = fct_reorder(HLA, number, .desc = FALSE))
```

Plt2: how many Peptides percentage (number of peptides over all peptides?) per HLA for positive T-cell responses

Here: amount of peptides out of all positive T-cell responses

```{r}
experiment_data |>
  filter(response == 1)|>
    group_by(HLA)|>
  summarize(number = n_distinct(Peptide))|>
  ungroup()|>
  mutate(
    total_peptides = sum(number),
    percentage_peptides = number/total_peptides * 100
  )
  
```

Here amount of peptides out of all peptides:

```{r}
total_peptides <- experiment_data|>
  summarize(total = n_distinct(Peptide)) |> 
  pull(total)
  
experiment_data |>
  filter(response == 1)|>
  group_by(HLA)|>
  summarize(number = n_distinct(Peptide))|>
  ungroup()|>
  mutate(
    percentage_peptides = number/total_peptides * 100
  )
```

```{r}
preprocess_hla <- function(data, value_col = "number", normalize = FALSE){
  
  preprocessed_data <- data |>
    mutate(HLA = str_remove(HLA, ":"))|>
    filter(response == 1)|>
    group_by(HLA)|>
    summarize(number = n_distinct(Peptide))|>
    ungroup()|>
    arrange(desc(HLA)) |> 
    mutate(HLA = fct_inorder(HLA))
  
  if (normalize){
    total_peptides <- data |>
      summarize(total = n_distinct(Peptide))|>
      pull(total)
    
    preprocessed_data <- preprocessed_data |>
      mutate(percentage_peptides = number/total_peptides*100)
  }
  
  return(preprocessed_data)
  
}
```

```{r}
plot_hla <- function(preprocessed_data, x_col, caption, show_y_text = TRUE){
  
  preprocessed_data |>
    ggplot(
    aes(x= !!sym(x_col), 
        y = HLA)
  ) + 
  geom_col(aes(width = 0.5), 
           fill = "darkblue")+
  geom_text(aes(label = if (x_col == "percentage_peptides") round(percentage_peptides, 1) else number),
              hjust = -0.2) +
  coord_cartesian(clip = "off")+
    #expand x-scale for labels on bars by 20%
  scale_x_continuous(expand = expansion(mult = c(0,.2)))+
  labs(caption = caption)+
  theme(aspect.ratio = 2/1, 
        axis.title.x = element_blank(), 
        axis.title.y = element_blank(), 
        axis.text.y = if (show_y_text) element_text(color = "black") else element_blank(),
        axis.ticks = element_line(color = "black"),
        axis.line = element_line(color = "black"),    
        panel.background = element_rect(fill = "white"),
    plot.caption = element_text(size = 12)
  )
  
}
```

```{r}
pl1 <- preprocess_hla(experiment_data)|>
  plot_hla(x_col = "number",
           caption = "Number of HLA-specific\nSARS-CoV-2 epitopes",
           show_y_text = TRUE)

pl2 <- preprocess_hla(experiment_data, normalize = TRUE)|>
  plot_hla(x_col = "percentage_peptides",
           caption="Immunogenic peptides\n(%, normalized)",
           show_y_text = FALSE)

combined <- pl1 | pl2 +
  plot_layout(widths = c())

combined
  
```

all in one call

```{r}

total_peptides <- experiment_data|>
  summarize(total = n_distinct(Peptide)) |> 
  pull(total)




pl1 <- experiment_data |>
  mutate(HLA = str_remove(HLA, ":"))|>
  filter(response == 1)|>
  group_by(HLA)|>
  summarize(number = n_distinct(Peptide))|>
  ungroup()|>
  arrange(desc(HLA)) |> 
  #fixes factor levels in exactly the row order appearing now
  mutate(HLA = fct_inorder(HLA))|>
  
  
  ggplot(
    aes(x= number, 
        y = HLA)
  ) + 
  geom_col(aes(width = 0.5), 
           fill = "darkblue")+
  geom_text(aes(label = number), 
            hjust = -0.2)+
  coord_cartesian(clip = "off")+
  scale_x_continuous(expand = expansion(mult = c(0,.2)))+
  labs(caption = "Number of HLA-specific\nSARS-CoV-2 epitopes")+
  theme(aspect.ratio = 2/1, 
        axis.title.x = element_blank(), 
        axis.title.y = element_blank(), 
        axis.ticks = element_line(color = "black"),
        axis.line = element_line(color = "black"),    
        panel.background = element_rect(fill = "white"),
    plot.caption = element_text(size = 12)
  )


  
pl2 <- experiment_data |>
  filter(response == 1)|>
  group_by(HLA)|>
  summarize(number = n_distinct(Peptide))|>
  ungroup()|>
  mutate(
    percentage_peptides = number/total_peptides * 100
  )|>

   ggplot(
    aes(x= percentage_peptides, 
        y = HLA)
  ) + 
  geom_col(aes(width = 0.5), 
           fill = "darkblue")+
  geom_text(aes(label = round(percentage_peptides, digits = 1)), 
            hjust = -0.2)+
  coord_cartesian(clip = "off")+
  scale_x_continuous(expand = expansion(mult = c(0,.2)))+
  labs(caption = "Immunogenic peptides\n (%, normalized)")+
  theme(aspect.ratio = 2/1, 
        axis.title.x = element_blank(), 
        axis.title.y = element_blank(), 
        axis.text.y  = element_blank(),
        axis.ticks = element_line(color = "black"),
        axis.line = element_line(color = "black"),    
        panel.background = element_rect(fill = "white"),
    plot.caption = element_text(size = 12)
  )
   
   
   
   

combined <- pl1 | pl2 +
  plot_layout(widths = c())

combined
```

# Plot 1G

```{r}

```

# Plot 1H

```{r}
experiment_data |>
  filter(Protein == "E")
```

```{r}
experiment_data|>
  mutate(Protein = toupper(Protein))|>
    group_by(Protein)|>
    summarize(number_Tcells = sum(response == 1))
```

```{r}
preprocess_protein <- function(data, value_col = "number_Tcells", normalize = FALSE){
  
  preprocessed_data <- data |>
    mutate(Protein = toupper(Protein))|>
    group_by(Protein)|>
    summarize(number_Tcells = sum(response == 1))|>
    ungroup()|>
    arrange((Protein)) |> 
    mutate(Protein = fct_inorder(Protein))
  
  if (normalize){
    total_peptides <- data |>
      summarize(total = n_distinct(Peptide))|>
      pull(total)
    
    preprocessed_data <- preprocessed_data |>
      mutate(percentage_peptides = number_Tcells/total_peptides*100)
  }
  
  return(preprocessed_data)
  
}

```

```{r}
plot_protein <- function(preprocessed_data, x_col, caption, show_y_text = TRUE){
  
  preprocessed_data |>
    ggplot(
    aes(x= !!sym(x_col), #evaluate variable string as a column name 
        y = Protein,
        fill = Protein)
  ) + 
  geom_col(aes(width = 0.5))+
  geom_text(aes(label = if (x_col == "percentage_peptides") round(percentage_peptides, 1) else number_Tcells),
              hjust = -0.2) +
  coord_cartesian(clip = "off")+
    #expand x-scale for labels on bars by 20%
  scale_x_continuous(expand = expansion(mult = c(0,.2)))+
  labs(caption = caption)+
  theme(aspect.ratio = 2/1, 
        axis.title.x = element_blank(), 
        axis.title.y = element_blank(), 
        axis.text.y = if (show_y_text) element_text(color = "black") else element_blank(),
        axis.ticks = element_line(color = "black"),
        axis.line = element_line(color = "black"),    
        panel.background = element_rect(fill = "white"),
    plot.caption = element_text(size = 12),
    legend.position = "none"
    )
  
}
```

```{r}
pl1 <- preprocess_protein(experiment_data)|>
  plot_protein(x_col = "number_Tcells",
           caption = "SARS-CoV-2\nT cell responses",
           show_y_text = TRUE)

pl2 <- preprocess_protein(experiment_data, normalize = TRUE)|>
  plot_protein(x_col = "percentage_peptides",
           caption="Immunogenic peptides\n(%, normalized)",
           show_y_text = FALSE)

combined <- pl1 | pl2 +
  plot_layout(widths = c())

combined
```

# Plot 1I

```{r}

```

# Plot 2A - funny plot

```{r}

```
