---
title: "05_analysis_1"
format: html
---

```{r}
# First main analysis
# e.g. immunodominance landscape similar to Fig. 1E
```

## Load the libraries

```{r}
library(dplyr)
library(ggplot2)
library(ggbreak)
library(ggforce)

```

## Graph

```{r}
complete_data |> 
  colnames()
```

```{r}
#goodone
library(dplyr)
library(ggplot2)
library(ggbreak)
library(ggforce)

#we need to do it in clean
complete_data <- complete_data |>
  mutate(Protein = str_to_upper(Protein))
  

protein_colors <- c(
  "ORF1" = "#1f77b4",
  "S" = "#ff7f0e",  
   "ORF3"= "#2ca02c",
  "ORF6" = "pink",
  "ORF7" = "red",
  "ORF8" = "lightblue",
  "ORF10" = "yellow",
  "E" = "purple",
  "M" = "darkblue",
  "N" = "darkred"
)

extras <- complete_data |>
  mutate(
    point_color = if_else(response == 0, "grey",protein_colors [Protein]),     
    point_fill  = if_else(response == 0, "grey", protein_colors [Protein]),
    HLA = factor(HLA, levels = c("A01:01","A02:01","A03:01","A24:02", "B07:02", "B08:01", "B15:01")))
  
  
  extras |> 
    ggplot(aes(
    x = `HLA-peptide`,
    y = log_fold_change,
    colour = point_color,
    fill   = point_fill,
    size = estimatedFrequency
  )) +
  geom_point() +

  facet_grid(~ HLA, scales = "free_x", space = "free") +
  geom_hline(yintercept = 2, linetype = "dashed") +
  scale_color_identity(guide = "none") +   
  scale_fill_identity(guide = "none") +
  scale_y_continuous(limits = c(0, NA), breaks = seq(0, 14, by = 2)) +
  scale_y_break(breaks = c(0, 2)) +
 theme(
  panel.spacing = unit(0, "lines"),  
  panel.border = element_rect(colour = "grey", fill = NA, size = 0.3),
  strip.text = element_text(angle = 90, hjust = 0.5, size = 10, face = "bold"),
  legend.position = "bottom",
   axis.text.y.right = element_blank(),
  axis.ticks.y.right = element_blank(),
  strip.background = element_blank(),
  axis.text.x = element_blank(),
  axis.text.y.left = element_text(size = 12),
  title = element_text(size = 16),
    legend.title = element_text(size = 10)

   ) +

  labs(title = "HLA",
       x = "Peptide HLA pairs",
       y = "Log fold change",
       size = "Estimated frequency"
) 

```

```{r}
#goodone
library(dplyr)
library(ggplot2)
library(ggbreak)
library(ggforce)

#we need to do it in clean
complete_data <- complete_data |>
  mutate(Protein = str_to_upper(Protein))
  

protein_colors <- c(
  "ORF1" = "#1f77b4",
  "S" = "#ff7f0e",  
   "ORF3"= "#2ca02c",
  "ORF6" = "pink",
  "ORF7" = "red",
  "ORF8" = "lightblue",
  "ORF10" = "yellow",
  "E" = "purple",
  "M" = "darkblue",
  "N" = "darkred"
)

extras <- complete_data |>
  mutate(
    point_color = if_else(response == 0, "grey",protein_colors [Protein]),     
    point_fill  = if_else(response == 0, "grey", protein_colors [Protein]),
    HLA = factor(HLA, levels = c("ORF1","S","ORF3","N", "M", "ORF8", "ORF7", "ORF6","ORF10", "E")))
  
  
  extras |> 
    ggplot(aes(
    x = `HLA-peptide`,
    y = log_fold_change,
    colour = point_color,
    fill   = point_fill,
    size = estimatedFrequency
  )) +
  geom_point() +

  facet_grid(~ Protein, scales = "free_x", space = "free") +
  geom_hline(yintercept = 2, linetype = "dashed") +
  scale_color_identity(guide = "none") +   
  scale_fill_identity(guide = "none") +
  scale_y_continuous(limits = c(0, NA), breaks = seq(0, 14, by = 2)) +
  scale_y_break(breaks = c(0, 2)) +
 theme(
  panel.spacing = unit(0, "lines"),  
  panel.border = element_rect(colour = "grey", fill = NA, size = 0.3),
  strip.text = element_text(angle = 90, hjust = 0.5, size = 10, face = "bold"),
  legend.position = "bottom",
   axis.text.y.right = element_blank(),
  axis.ticks.y.right = element_blank(),
  strip.background = element_blank(),
  axis.text.x = element_blank(),
  axis.text.y.left = element_text(size = 12),
  title = element_text(size = 16),
    legend.title = element_text(size = 10)

   ) +

  labs(title = "Protein",
       x = "Peptide HLA pairs",
       y = "Log fold change",
       size = "Estimated frequency"
) 
```{r}
#| warning: true
library("tidyverse")
library("here")
library("paletteer") #remove this package
library("forcats") #same
```

## Figure 1I - frequency of epitopes across proteins of origin

It is interesting to assess how different genes from SARS-CoV-2 genome contribute in generating immunogenic epitopes. To do this, we will plot the total frequency of T cell responses for each patient, distributed across the whole genome.

```{r}
#| title: "total estimated frequency per patient"

path = here("data", "03_data_augment.tsv")
augmented_data <- read_tsv(file = path, show_col_types = FALSE)

augmented_data <- augmented_data |>
  mutate ( estimatedFrequency = estimatedFrequency*100, #converts into percentage
           total_estimated_frequency = total_estimated_frequency*100,
  sampleID = fct_reorder(sampleID, total_estimated_frequency)) |> #reorders based on value of second argument
  arrange(desc(total_estimated_frequency))

```

```{r}
#| out-width: 100%
#| fig-width: 10
palette10 <- c(
  "#E69F00", "#56B4E9", "#009E73","#00AFAA", "#0072B2","#D55E00", "#CC79A7", "#F0E442","#0099BB", "#5A61FF")

ggplot(augmented_data,
       aes(x = estimatedFrequency, y = sampleID, fill = Protein)) +
  geom_col( position = "stack") +
  scale_fill_manual(values  = palette10) +
  scale_x_continuous(breaks= c(0, 5, 10, 15, 20, 25)) +
  labs( x = "SARS-CoV-2 reactive CD8 T cells (%)", y = "Samples", title = "Frequency of SARS-CoV-2 reactive CD8 T cells in patients with COVID19, distributed across protein of origin")
```

As shown here, for most samples, the majority of reactive T cells recognize antigens derived from ORF1. Taking into account the size of the genes, hence the amount of peptides that can be generated, it is reasonable to witness a bigger response against bigger genes (such as ORF1). However, we can perform some normalization to calculate the percentage of immunogenic peptides that arise from each gene. (Selina's figures).
Libraries

```{r}
library("here")
library("tidyverse")
```

```{r}
source("99_proj_func.R")
```

Load data

```{r}
experiment_data <- load_tsvData("03_data_augment.tsv", show_col_types = FALSE)
```

# Plot 1E

```{r}

```

# Plot 1F

```{r}

```

# Plot 1G

```{r}

```

# Plot 1H

```{r}

```

# Plot 1I

```{r}

```

# Plot 2A - funny plot

```{r}

```
