---
title: "05_analysis_1"
format: html
---

```{r}
# First main analysis
# e.g. immunodominance landscape similar to Fig. 1E
```

```{r}
#| warning: true
library("tidyverse")
library("here")
library("paletteer") #remove this package
library("forcats") #same
```

## Figure 1I - frequency of epitopes across proteins of origin

It is interesting to assess how different genes from SARS-CoV-2 genome contribute in generating immunogenic epitopes. To do this, we will plot the total frequency of T cell responses for each patient, distributed across the whole genome.

```{r}
#| title: "total estimated frequency per patient"

path = here("data", "03_data_augment.tsv")
augmented_data <- read_tsv(file = path, show_col_types = FALSE)

augmented_data <- augmented_data |>
  mutate ( estimatedFrequency = estimatedFrequency*100, #converts into percentage
           total_estimated_frequency = total_estimated_frequency*100,
  sampleID = fct_reorder(sampleID, total_estimated_frequency)) |> #reorders based on value of second argument
  arrange(desc(total_estimated_frequency))

```

```{r}
#| out-width: 100%
#| fig-width: 10
palette10 <- c(
  "#E69F00", "#56B4E9", "#009E73","#00AFAA", "#0072B2","#D55E00", "#CC79A7", "#F0E442","#0099BB", "#5A61FF")

ggplot(augmented_data,
       aes(x = estimatedFrequency, y = sampleID, fill = Protein)) +
  geom_col( position = "stack") +
  scale_fill_manual(values  = palette10) +
  scale_x_continuous(breaks= c(0, 5, 10, 15, 20, 25)) +
  labs( x = "SARS-CoV-2 reactive CD8 T cells (%)", y = "Samples", title = "Frequency of SARS-CoV-2 reactive CD8 T cells in patients with COVID19, distributed across protein of origin")
```

As shown here, for most samples, the majority of reactive T cells recognize antigens derived from ORF1. Taking into account the size of the genes, hence the amount of peptides that can be generated, it is reasonable to witness a bigger response against bigger genes (such as ORF1). However, we can perform some normalization to calculate the percentage of immunogenic peptides that arise from each gene. (Selina's figures).
Libraries

```{r}
library("here")
library("tidyverse")
```

```{r}
source("99_proj_func.R")
```

Load data

```{r}
experiment_data <- load_tsvData("03_data_augment.tsv", show_col_types = FALSE)
```

# Plot 1E

```{r}

```

# Plot 1F

```{r}

```

# Plot 1G

```{r}

```

# Plot 1H

```{r}

```

# Plot 1I

```{r}

```

# Plot 2A - funny plot

```{r}

```
